---
title: "A Tutorial for Covariate Adjustment for Binary outcomes in Clinical Trials"
author: ''
date: "`r format(Sys.time(), '%Y-%m-%d %I:%M')`"
output:
  html_document:
    toc: true # table of content true
    # toc_depth: 3
    number_sections: true
    theme: united
    highlight: tango
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

This tutorial intoduces how to perform covariate-adjusted analysis for binary oucomes in randomized clinical trials. 

In the second section, we provide an example data.

In Section 3, we give a few estimands for binary outcomes.

In Section 4, we introduce two covariate-adjusted estimators and demonstrate how to implement them in `R`.

Before we start, we load the following packages.

```{r, message=F}
library(tidyverse) # for data cleaning and manipulation
library(boot) # for bootstrap
library(devtools) # optional, for sourcing R script from GitHub.
```

# Example data: Buprenorphine tapering and illicit opioid use  (NIDA-CTN-0003)

The trial of ``Buprenorphine tapering schedule and illicit opioid use''  in the National Drug Abuse Treatment Clinical Trials Network (NIDA-CTN-0003), is a phase-3 randomized trial completed in 2005 [@CTN03]. The goal was to compare the effects of a short or long taper schedule after buprenorphine stabilization of patients with opioid use disorder. Patients were randomized into two arms: 28-day taper (control, 259 patients, 36\% missing outcomes) and 7-day taper (treatment, 252 patients, 21\% missing outcomes), stratified by maintenance dose (3 levels) measured at randomization. The outcome of interest is a binary indicator of whether a participant's urine tested at the end of the study is opioid-free. In addition to the stratification variable, we adjust for the following baseline variables: sex, opioid urine toxicology results, the Adjective Rating Scale for Withdrawal (ARSW), the Clinical Opiate Withdrawal Scale (COWS) and the Visual Analog Scale (VAS).

_Note: The above description is from [@wang2019model]. The data for this example are not publicly available, but they can be requested from the NIDA databases at https://datashare.nida.nih.gov. _

## Reading in data

```{r, eval=FALSE}
CTN03 <- readRDS("data/CTN03.rds")
head(CTN03)
```

```{r, echo=F}
setwd("~/Dropbox/JHU/clinical-trial/covariate-adaptive/Covariate-adaptive/")
CTN03 <- readRDS("data/CTN03.rds")[,c(4,2,5,3,6,7,8,9)]
head(CTN03)
```

In the data frame `CTN03`, the outcome is 

  - `outcome.OPI`: a binary indicator of whether a participant's urine tested at the end of the study is opioid-free.
  
The treatment allocation variable is 

  - `arm`: "7-day taper" (treatment) or "28-day taper" (control).
  
The baseline variables are

  - `baseline.OPI`: baseline opioid urine toxicology results,
  
  - `sex`: "Female" or "Male",
  
  - `COWS`: Clinical Opiate Withdrawal Scale,
  
  - `ARSW`: Adjective Rating Scale for Withdrawal,
  
  - `VAS`:  Visual Analog Scale,
  
  - `strata`: maintenance dose (3 levels, 8, 16 or 24).

## Preprocessing

The preprocessing steps for `CTN03` include:

  - Converting treatment variable `arm` from character to binary numbers (1 for treatment and 0 for control).
  
  - Converting the variable type of `strata` from "numerical"" to "character" and removing rows with missing strata.
  
  - Imputing all missing baseline values by the median for continuous variables and mode for binary or categorical variables. 
  
The corresponding R code is given below:

```{r}
# covert arm variable to binary numbers
CTN03$arm <- as.numeric(CTN03$arm == "7-day taper")
# covert strata variable to factor
CTN03$strata <- as.factor(CTN03$strata)
# impute missing baseline values
CTN03$baseline.OPI[which(is.na(CTN03$baseline.OPI))] <- median(CTN03$baseline.OPI, na.rm = T)
CTN03$COWS[which(is.na(CTN03$COWS))] <- median(CTN03$COWS, na.rm = T)
CTN03$ARSW[which(is.na(CTN03$ARSW))] <- median(CTN03$ARSW, na.rm = T)
CTN03$VAS[which(is.na(CTN03$VAS))] <- median(CTN03$VAS, na.rm = T)
CTN03 <- CTN03[!is.na(CTN03$strata),]
```


# Estimands

Let $Y$ denote the binary outcome variable, $A$ denote the binary treatment indicator ($A=1$ for treatment and $A=0$ for control) and $X$ denote a vector of baseline variables. The parameters of interest include:

 - The average treatment effect (ATE, or risk difference), defined as $\Delta = P(Y=1|A=1) - P(Y=0|A=1)$.
 
 - Relative risk (RR), defined as $\Gamma = P(Y=1|A=1) / P(Y=0|A=1)$.
 
# Estimators

Let $n$ be the sample size and $(Y_i, A_i, X_i)$ be the observed data vector for participant $i, i = 1,\dots, n$.

## The standardized logistic regression estimator (assuming no missing outcomes)

### Point estimates

The standardized logistic regression estimator is calculated by first fitting a working model: $$P(Y=1|A,X) = \mbox{expit}(\beta_0 + \beta_A A + \beta_{X}^t X),$$
where $\mbox{expit}(x) = 1/(1+e^{-x})$, and getting the maximum likelihood estimates $(\widehat{\beta}_0, \widehat{\beta}_A, \widehat{\beta}_{X}^t)^t$. Then define
$$\widehat{\Delta}_{logistic} = \frac{1}{n}\sum_{i=1}^n\left\{\mbox{expit}(\widehat\beta_0 + \widehat\beta_A + \widehat\beta_{X}^tX_i) - \mbox{expit}(\widehat\beta_0 + \widehat\beta_{X}^t X_i)\right\} $$ and $$\widehat{\Gamma}_{logistic} = \frac{\sum_{i=1}^n\mbox{expit}(\widehat\beta_0 + \widehat\beta_A + \widehat\beta_{X}^tX_i)}{\sum_{i=1}^n\mbox{expit}(\widehat\beta_0 + \widehat\beta_{X}^tX_i)}. $$

The following code can be used to calculate $\widehat{\Delta}_{logistic}$ and $\widehat{\Gamma}_{logistic}$. For a different data set, `outcome.OPI` below should be replaced by the outcome variable name and `CTN03` below should be replaced by the name of the data.

```{r}
logistic_fit <- glm(outcome.OPI~., data = CTN03, family = "binomial")
p1 <- predict(logistic_fit, mutate(CTN03, arm = 1), type = "response")
p0 <- predict(logistic_fit, mutate(CTN03, arm = 0), type = "response")
ATE_logistic <- mean(p1) - mean(p0)
RR_logistic <- mean(p1) / mean(p0)
c(ATE = ATE_logistic, RR = RR_logistic)
```

### Inference under simple randomization

If the clinical trial is done by simple randomization, then the variances and confidence intervals of $\widehat{\Delta}_{logistic}$ and $\widehat{\Gamma}_{logistic}$ can be approximated by bootstrap using the `boot` package in `R`, as demonstrated by the following R code:

```{r}
R <- 2000 # number of replicates for bootstrap
# Inference for ATE
logistic_ATE <- function(d, i){
  logistic_fit <- glm(outcome.OPI~., data = d[i,], family = "binomial")
  p1 <- predict(logistic_fit, mutate(d, arm = 1), type = "response")
  p0 <- predict(logistic_fit, mutate(d, arm = 0), type = "response")
  mean(p1) - mean(p0)
}
boot_ATE <- boot(CTN03, statistic = logistic_ATE, R = R)
# Inference for RR
logistic_RR <- function(d, i){
  logistic_fit <- glm(outcome.OPI~., data = d[i,], family = "binomial")
  p1 <- predict(logistic_fit, mutate(d, arm = 1), type = "response")
  p0 <- predict(logistic_fit, mutate(d, arm = 0), type = "response")
  mean(p1)/mean(p0)
}
boot_RR <- boot(CTN03, statistic = logistic_RR, R = R)
#summarizing result
output <- matrix(NA, nrow = 2, ncol = 4, dimnames = list(c("ATE", "RR"), c("est", "var", "CI.lower", "CI.upper")))
output[1,] <- c(boot_ATE$t0, var(boot_ATE$t), boot.ci(boot_ATE, type = "bca")$bca[4:5])
output[2,] <- c(boot_RR$t0, var(boot_RR$t), boot.ci(boot_RR, type = "bca")$bca[4:5])
round(output, 3)
```

### Inference under stratified or biased-coin randomization

If the clinical trial is done by stratified or biased-coin randomization, the variance and confidence intervals obtained by the above bootstrap can be conservative [@wang2019model]. One way to avoid conservativeness is to use the `ICAD.R` function for inference (available at https://github.com/BingkaiWang/covariate-adaptive/blob/master/R/ICAD.R). 

The `ICAD.R` function takes input of $Y$ (outcome), $A$ (treatment), $X$ (baseline variables), $\pi$ (targeted $P(A=1)$) and family ("binomial" for binary outcomes). The `ICAD.R` function outputs a summary table gives point estimats, variance and 95\% confidence intervals for unadjusted and adjusted estimators. The following code chunk demonstrates how to use it.
 
```{r, warning = F}
source_url("https://raw.githubusercontent.com/BingkaiWang/covariate-adaptive/master/R/ICAD.R")
Y <- CTN03$outcome.OPI
A <- CTN03$arm
X <- CTN03[, 3:7]
Strata <- CTN03$strata
ICAD(Y, A, Strata, X, pi = 0.5, family = "binomial") %>% round(3)
```

## The DR-WLS estimator (handling missing outcomes under the missing at random assumption)

### Point estimates

The doubly robust weighted least squared (DR-WLS) estimator from Marshall Joffe is described by [@robins2007]. This estimator involves two working models: an outcome regression model for the outcome and a propensity score model for missing outcomes. In both models, baseline variables in $X$ are adjusted. This estimator is consistent as long as at least one of the models is correct, under the missing at random assumption and the assumption that everyone has a positive probability of completing the trial (i.e., having their primary outcome measured) regardless of their baseline variables. 

Similar to the standardized logistic regression estimator, statistical inference for the DR-WLS estimator can be done by bootstrap under simple randomization and can be outputted by the `ICAD.R` function under stratified or biased-coin randomization.

```{r, warning=F}
M <- !is.na(CTN03$outcome.OPI) # M is an indicator of non-missingness
propensity_fit <- glm(M ~., data = cbind(M, CTN03[,2:8]), family = "binomial")
propensity_score <- predict(propensity_fit, type = "response")
drwls_fit <- glm(outcome.OPI ~., data = CTN03, family = "binomial", weights = 1/propensity_score)
p1 <- predict(drwls_fit, mutate(CTN03, arm = 1), type = "response")
p0 <- predict(drwls_fit, mutate(CTN03, arm = 0), type = "response")
ATE_drwls <- mean(p1) - mean(p0)
RR_drwls <- mean(p1) / mean(p0)
c(ATE = ATE_drwls, RR = RR_drwls)
```

### Inference under simple randomization

If the clinical trial is done by simple randomization, then the variances and confidence intervals of $\widehat{\Delta}_{logistic}$ and $\widehat{\Gamma}_{logistic}$ can be approximated by bootstrap using the `boot` package in `R`, as demonstrated by the following R code:

```{r, warning=F}
R <- 2000 # number of replicates for bootstrap
# Inference for ATE
drwls_ATE <- function(d, i){
  d <- d[i,]
  M <- !is.na(d$outcome.OPI) # M is an indicator of non-missingness
  propensity_fit <- glm(M ~., data = cbind(M, d[,2:8]), family = "binomial")
  propensity_score <- predict(propensity_fit, type = "response")
  drwls_fit <- glm(outcome.OPI ~., data = d, family = "binomial", weights = 1/propensity_score)
  p1 <- predict(drwls_fit, mutate(d, arm = 1), type = "response")
  p0 <- predict(drwls_fit, mutate(d, arm = 0), type = "response")
  mean(p1) - mean(p0)
}
boot_ATE <- boot(CTN03, statistic = drwls_ATE, R = R)
# Inference for RR
drwls_RR <- function(d, i){
  d <- d[i,]
  M <- !is.na(d$outcome.OPI) # M is an indicator of non-missingness
  propensity_fit <- glm(M ~., data = cbind(M, d[,2:8]), family = "binomial")
  propensity_score <- predict(propensity_fit, type = "response")
  drwls_fit <- glm(outcome.OPI ~., data = d, family = "binomial", weights = 1/propensity_score)
  p1 <- predict(drwls_fit, mutate(d, arm = 1), type = "response")
  p0 <- predict(drwls_fit, mutate(d, arm = 0), type = "response")
  mean(p1) / mean(p0)
}
boot_RR <- boot(CTN03, statistic = drwls_RR, R = R)
#summarizing result
output <- matrix(NA, nrow = 2, ncol = 4, dimnames = list(c("ATE", "RR"), c("est", "var", "CI.lower", "CI.upper")))
output[1,] <- c(boot_ATE$t0, var(boot_ATE$t), boot.ci(boot_ATE, type = "bca")$bca[4:5])
output[2,] <- c(boot_RR$t0, var(boot_RR$t), boot.ci(boot_RR, type = "bca")$bca[4:5])
round(output, 3)
```

### Inference under stratified or biased-coin randomization

Similar to the standardized logistic regression estimator, under stratified or biased-coin randomization, the variance and confidence intervals of the DR-WLS estimator for the average treatment effect can be obtained by the `ICAD.R` function.

The R code is the same as in Section 4.1.3, where the result for the DR-WLS estimator is given in the last row of the output.

# References